---
output: github_document
---

<!-- badges: start -->
  [![Codecov test coverage](https://codecov.io/gh/HanLabCollaboration/ReproducibleYostIndex/graph/badge.svg)](https://app.codecov.io/gh/HanLabCollaboration/ReproducibleYostIndex)
  <!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
```

# ReproducibleYostIndex

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
The goal of `ReproducibleYostIndex` is to provide a simple, tidy interface for calculating the **Yost Index**, a composite measure of neighborhood-level socioeconomic status (SES), using data from the US Census Bureau's American Community Survey (ACS).

## What is the Yost Index?

The Yost Index is a score calculated using factor analysis on 7 key variables related to socioeconomic deprivation. This package automatically fetches the required ACS data and computes the following component variables:

* **income**: Median Household Income
* **wkcls**: Percent working class
* **unemp**: Percent unemployed
* **educ**: Education score
* **poverty150**: Percent below 150% poverty line
* **rent**: Median Gross Rent
* **hval**: Median Home Value

## Features

* **Easy Data Fetching**: Wraps `tidycensus` to get all required variables automatically.
* **Flexible Geographies**: Compute the index for "county", "tract", or "block group" levels.
* **Variable Scopes**: Calculate the index using a "national" scope (ranking all geographies against each other) or a "state" scope (ranking geographies within each state).
* **Imputation**: Includes an option for simple spatial imputation for missing data (`impute = TRUE`).

## Installation

You can install the development version of ReproducibleYostIndex from [GitHub](https://github.com/HanLabCollaboration/ReproducibleYostIndex) with:

``` r
# install.packages("pak")
# pak::pak("HanLabCollaboration/ReproducibleYostIndex")
```

## Example

To use `ReproducibleYostIndex`, you will also need a **Census API Key**. You can get one [here](https://api.census.gov/data/key_signup.html) and install it using `tidycensus::census_api_key("YOUR_KEY_HERE", install = TRUE)`.

This example computes the Yost Index for all counties in California for the year 2022.

```{r example, message=FALSE, warning=FALSE}
library(ReproducibleYostIndex)
library(dplyr)

# Compute the Yost Index for California counties
# This uses scope = "national" by default, comparing CA counties to all US counties
yost_ca_county <- computeYostIndex(
  geo = "county",
  year = 2022,
  states = "CA",
  impute = FALSE,
  quiet = TRUE,
  scope = "state",
)

# The result is a list
names(yost_ca_county)

# The main data frame is $df_yost
glimpse(yost_ca_county$df_yost)

# You can quickly see the distribution of Yost Quintiles
yost_ca_county$df_yost %>% 
  count(YostQuintile, sort = TRUE)

# You can also inspect the factor analysis object
# yost_ca_county$obj_factor
```

## License

This package is available under the MIT License.
