---
title: "ReproduceYostIndex Walkthrough"
author: " Rafael Catoia Pulgrossi | [QSU](https://med.stanford.edu/qsu 'View website') | [Tae Yoon Lee](https://profiles.stanford.edu/tae-yoon-lee 'View website') | [Summer Han](https://med.stanford.edu/summerhanlab.html 'View website')"
date: today
format:
  html:
    self-contained: true
    theme: united
    toc: true
    toc-expand: 4
    toc-depth: 4
    code-fold: true
    toc-title: 'Contents'
    page-layout: full
    grid:
      body-width: 1440px
      gutter-width: 1.5rem
execute:
  freeze: true
editor_options:
  chunk_output_type: console
---

```{r}
#| include: false
library(ReproduceYostIndex)
library(ggplot2)
library(dplyr)
library(sf)
```

## Two Ways to Get Yost Index Data

`ReproduceYostIndex` provides two functions that serve different needs:

| | `getYostIndex()` | `computeYostIndex()` |
|---|---|---|
| **Census API key** | Not required | Required |
| **Speed** | Instant (pre-computed) | Minutes to hours |
| **Geographies** | County, tract, block group | County, tract, block group, state |
| **Scope** | National, state | National, state, county |
| **Variants returned** | Four YostIndex possible versions (Raw/Stabilized/Imputed/Stabilized+Imputed) | One (your chosen parameters) |
| **Best for** | Most analyses, quick exploration | Custom pipelines |

**Start with `getYostIndex()`**. It downloads pre-computed files from GitHub — no Census API key, no wait, all four Yost variants in a single call. Switch to `computeYostIndex()` only when you need a year not yet pre-computed, a county-level scope, or a fully custom pipeline.

---

## `getYostIndex()` — Recommended Starting Point

### Installation

```r
# install.packages("devtools")
devtools::install_github("HanLabCollaboration/ReproduceYostIndex")
```

### Basic usage

```{r getyostindex_basic}
#| eval: true
#| message: false

library(ReproduceYostIndex)

# All US counties, 2022 — no Census API key needed
county_yost <- getYostIndex(geo = "county", year = 2022)
county_yost |>  head(10) |> DT::datatable()
```

A single call returns **four Yost Index options** side by side:

| Column | Description |
|---|---|
| `Yost` / `YostQuintile` | Baseline — no stabilization, no imputation |
| `YostStabilized` / `YostStabilizedQuintile` | Stabilization applied |
| `YostImputed` / `YostImputedQuintile` | Spatial imputation for missing values |
| `YostStabilizedImputed` / `YostStabilizedImputedQuintile` | Both stabilization and imputation |

### Filtering by state and scope

```{r getyostindex_filter}
#| eval: true
#| message: false

# Tracts for California and New York, ranked within each state
tracts_ca_ny <- getYostIndex(
  geo    = "tract",
  year   = 2022,
  states = c("CA", "NY"),
  scope  = "national"
)
glimpse(tracts_ca_ny)
```

### Choosing a scope

- **`"national"`** (default): harmonizes US geographies (state,counties,tracts or census block groups) against each other and performs only one factor analysis.
- **`"state"`**: harmonizes and performs factor analysis separately for each state.
- **`"county"`**: harmonizes and performs factor analysis separately for each county.


### Choosing a Yost variant

Pick the column that fits your analysis:

```{r getyostindex_variants}
#| eval: true
#| message: false

df <- getYostIndex(geo = "county", year = 2022, states = "CA")

# For most county/tract analyses: plain Yost
hist(df$Yost, main = "Yost (baseline)", xlab = "Yost score", col = "steelblue", border = "white")

# For block groups or small-sample tracts: stabilized variant
# df$YostStabilized  — shrinks unreliable estimates toward their parent geography

# When there are many NAs: imputed variant
# df$YostImputed     — fills missing values via spatial mean of neighbors
```

### Caching

Files are downloaded once and cached locally. Subsequent calls with the same arguments return instantly:

```{r getyostindex_cache}
#| eval: false

# First call: downloads ~2 MB file
df <- getYostIndex(geo = "county", year = 2022)

# Second call: reads from cache — instantaneous
df <- getYostIndex(geo = "county", year = 2022)

# Force a fresh download
df <- getYostIndex(geo = "county", year = 2022, cache = FALSE)

# See where the cache lives
tools::R_user_dir("ReproduceYostIndex", "cache")
```

---

## `computeYostIndex()` — Full Pipeline

Use `computeYostIndex()` when:
- You need **county-level scope** (quintiles & factor within each county)
- You want to **inspect intermediate outputs** (raw ACS values, factor loadings, imputed data,the 7 yost variables)
- You are building a **custom pipeline** and need fine-grained control. You may choose to change from `rank` to `standardize`, or to use `average` instead of weighted average per `total_population`

### Census API key (required)

```r
# Get a free key at: https://api.census.gov/data/key_signup.html
tidycensus::census_api_key("YOUR_KEY_HERE", install = TRUE)
```

### Basic usage

```{r computeyost_basic}
#| eval: true
#| results: hide
#| message: false
#| warning: false

yost_ca <- computeYostIndex(
  geo       = "tract",
  year      = 2022,
  states    = "CA",
  scope     = "state",
  stabilize = TRUE,    # empirical Bayes stabilization
  impute    = TRUE,    # spatial imputation for missing values
  keep_geometry = TRUE
)
```

```{r computeyost_output}
#| eval: true

# Baseline (always present — no stabilization, no imputation)
head(yost_ca$df_yost_raw)

# Requested variant (stabilize = TRUE, impute = TRUE)
head(yost_ca$df_yost)
```

### Understanding the output list

```{r computeyost_structure}
#| eval: true

str(yost_ca, max.level = 1)
```

| Element | Contents |
|---|---|
| `df_yost_raw` | Baseline Yost — always `Yost` + `YostQuintile`, regardless of parameters |
| `df_yost` | Requested variant — column name reflects what was applied |
| `df_raw_values` | Raw ACS estimates and margins of error |
| `df_geometry` | Spatial geometries (if `keep_geometry = TRUE`) |
| `df_imputed` | Component variables after imputation |
| `df_rank` | Ranked/standardized inputs to factor analysis |
| `obj_factor` | `psych::fa` object(s) |

### Stabilization explained

At fine geographic levels (especially block groups), ACS margins of error can exceed the estimates themselves. The empirical Bayes stabilization method pulls each lower-level estimate toward its parent geography using a data-adaptive weight:

$$w = \frac{t^2}{s^2 + t^2}$$

- $t^2$: within-parent heterogeneity — how much the units in this parent actually differ
- $s^2 = (\text{MOE}/1.645)^2$: sampling variance from the ACS margin of error

When a unit is **reliable** ($s^2$ small), $w \approx 1$ — the estimate is used as-is.
When a unit is **noisy** ($s^2$ large), $w \approx 0$ — the estimate is pulled toward the parent.

```{r stabilize_compare}
#| eval: true
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 5

# Compare raw vs stabilized
comparison <- yost_ca$df_yost_raw |>
  dplyr::rename(Yost_raw = Yost) |>
  dplyr::left_join(
    yost_ca$df_yost |>
      dplyr::rename(Yost_stabilized = YostStabilizedImputed),
    by = c("GEOID", "annotation")
  )

ggplot(comparison, aes(x = Yost_raw, y = Yost_stabilized)) +
  geom_point(alpha = 0.15, size = 0.6, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red",
              linetype = "dashed", linewidth = 0.8) +
  labs(
    title    = "Raw vs. Stabilized Yost Index (CA tracts, 2022)",
    subtitle = "Points off the diagonal were pulled toward their parent geography",
    x        = "Yost (raw, no stabilization)",
    y        = "YostStabilizedImputed",
    caption  = "Red line = no change"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title    = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, color = "grey40"))
```

### Mapping with `computeYostIndex()`

Because `computeYostIndex()` can return geometry directly, mapping is straightforward:

```{r computeyost_map}
#| eval: true
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 5

df_map <- yost_ca$df_geometry |>
  dplyr::left_join(yost_ca$df_yost, by = "GEOID")

ggplot(df_map) +
  geom_sf(aes(fill = YostStabilizedImputed), color = "white", linewidth = 0.02) +
  scale_fill_viridis_c(option = "plasma", name = "Yost\n(Stabilized)") +
  labs(
    title    = "Neighborhood SES — California Census Tracts",
    subtitle = "YostStabilizedImputed | 2022 ACS | State scope",
    caption  = "Higher values = higher socioeconomic status"
  ) +
  theme_void() +
  theme(
    plot.title    = element_text(face = "bold", size = 15, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "grey40"),
    plot.caption  = element_text(hjust = 0.5, color = "grey60")
  )
```

### When to use `scope = "county"`

County-level scope is only available in `computeYostIndex()`. It ranks geographies within each county, which is useful for within-county resource allocation or studies confined to a single metro area:

```{r computeyost_county_scope}
#| eval: false

# Tracts in Los Angeles County ranked among themselves
yost_la <- computeYostIndex(
  geo    = "tract",
  year   = 2022,
  states = "CA",
  scope  = "county"   # not available in getYostIndex()
)
```

---

## Decision Guide

```{r decision_guide}
#| echo: false

knitr::kable(data.frame(
  Question = c(
    "Do I have a Census API key?",
    "Do I need data for 2024 or later?",
    "Do I need county-level scope?",
    "Do I need intermediate outputs (raw ACS, factor loadings)?",
    "Am I doing a quick analysis or teaching?"
  ),
  `If YES` = c(
    "Either function works",
    "Use computeYostIndex()",
    "Use computeYostIndex()",
    "Use computeYostIndex()",
    "Use getYostIndex()"
  ),
  `If NO` = c(
    "Use getYostIndex()",
    "Use getYostIndex()",
    "Use getYostIndex()",
    "Use getYostIndex()",
    "Use computeYostIndex() for full control"
  )
), col.names = c("Question", "If YES →", "If NO →"))
```

---

## Further Reading

- Yost K, et al. (2001). Socioeconomic status and breast cancer incidence in California. *Cancer Causes & Control*, 12(8), 703–711.
- Yang J, et al. (2014). Developing an area-based socioeconomic measure from ACS data. *Cancer Prevention Research*, 7(7).
- Package source: <https://github.com/HanLabCollaboration/ReproduceYostIndex>
